Automated Modelling of Flow in Porous Media via the Finite Element Method

### 2013-02-03 Papers/ABCM/abcm.tex
Add thesis stuff (Sat.Eq., B.C, MFEM, SUPG) in portuguese. Still missing translation, DOLFIN stuff, new results

%Paper for submission to ABCMs Journal
%Using the ENCIT template just for reference, still have to adapt to ABCM format(where is the Latex template for it?)
%Authors: Bruno G. B. Luna
%         Paulo R. M. Lyra
%         Ramiro B. Willmersdorf
%Date (Start): 02/12/2012
\documentclass[10pt,fleqn,a4paper]{article}
\usepackage{encit2012}
\usepackage{amsmath}
\usepackage[latin1]{inputenc}
\usepackage{calc,excludeonly,graphicx}
\usepackage[onehalfspacing]{setspace}
%\usepackage[color,notcite,notref]{showkeys}
\definecolor{DarkBlue}{rgb}{0,0,0.7} 
\usepackage[pdftex,colorlinks=true, linkcolor=DarkBlue, citecolor=DarkBlue]{hyperref}
\usepackage{threeparttable}

\begin{document}
\hspace{-8.5mm}
\begin{tabular}{||p{\textwidth}}
\begin{center}
\vspace{-4mm}
\title{Automated Modelling of Flow in Porous Media via the Finite Element Method  }
\end{center}
\authors{Bruno G. B. Luna, bruno\_bl@yahoo.com.br} \\
\authors{Paulo R. M. Lyra, prmlyra@padmec.org} \\
\authors{Ramiro B. Willmersdorf, ramiro@willmersdorf.net} \\
\institution{Federal University of Pernambuco (UFPE), Departament of Mechanical Engineering, Cidade Universit\'aria s/n, Recife, PE, Brazil, www.ufpe.br} \\
\\
\abstract{
The simulation of multiphase flows in porous media impose many challenges due to a series of factors as the high anisotropic and heterogeneous media handled in this type of analysis, the coupled elliptic-hyperbolic mathematical nature of the associated Partial Differential Equations (PDEs), the programming of the solution method, among others. This work presents the implementation of a software created using the FEniCs tool for automated low-level code generation and PyAMG for the fully automatic creation of consistent multilevel matrix structures used in Algebraic Multigrid Methods (AMG) to be applied to the numerical solution of porous media flows using the Finite Element Method (FEM). The methods described here are general enough to handle with three-dimensional, heterogeneous and anisotropic problems. Examples are shown and results discussed for two-dimensional problems in homogenous and heterogeneous domains with iso- and anisotropic permeability tensors.}\\
\\
\keywords{\textbf{Keywords:} Multigrid, Oil Reservoir, Porous Media Flow}\\
\end{tabular}

\section{NOMENCLATURE}

@@

\section{INTRODUCTION}

\section{MATHEMATICAL FORMULATION}

\subsection{Pressure Equation}

\subsection{Saturation Equation}

\subsection{Boundary and Initial Conditions}

\section{NUMERICAL FORMULATION}

\section{IMPLEMENTATION}

>>>>

%Paper for submission to ABCMs Journal
%Using the ENCIT template just for reference, still have to adapt to ABCM format(where is the Latex template for it?)
%Authors: Bruno G. B. Luna
%         Paulo R. M. Lyra
%         Ramiro B. Willmersdorf
%Date (Start): 02/12/2012

\documentclass[10pt,fleqn,a4paper]{article}
\usepackage{encit2012}
\usepackage{amsmath}
\usepackage[latin1]{inputenc}
\usepackage{calc,excludeonly,graphicx}
\usepackage[onehalfspacing]{setspace}
%\usepackage[color,notcite,notref]{showkeys}
\definecolor{DarkBlue}{rgb}{0,0,0.7} 
\usepackage[pdftex,colorlinks=true, linkcolor=DarkBlue, citecolor=DarkBlue]{hyperref}
\usepackage{threeparttable}

% Bruno: Custom packages
\usepackage{subfigure}
\newtheorem{problem}{Problem}
% Definition for tensors format
\def\mathbi#1{\textbf{\em #1}}

\begin{document}
\hspace{-8.5mm}
\begin{tabular}{||p{\textwidth}}
\begin{center}
\vspace{-4mm}
\title{Automated Modelling of Flow in Porous Media via the Finite Element Method  }
\end{center}
\authors{Bruno G. B. Luna, bruno\_bl@yahoo.com.br} \\
\authors{Paulo R. M. Lyra, prmlyra@padmec.org} \\
\authors{Ramiro B. Willmersdorf, ramiro@willmersdorf.net} \\
\institution{Federal University of Pernambuco (UFPE), Departament of Mechanical Engineering, Cidade Universit\'aria s/n, Recife, PE, Brazil, www.ufpe.br} \\
\\
\abstract{
The simulation of multiphase flows in porous media impose many challenges due to a series of factors as the high anisotropic and heterogeneous media handled in this type of analysis, the coupled elliptic-hyperbolic mathematical nature of the associated Partial Differential Equations (PDEs), the programming of the solution method, among others. This work presents the implementation of a software created using the FEniCs tool for automated low-level code generation and PyAMG for the fully automatic creation of consistent multilevel matrix structures used in Algebraic Multigrid Methods (AMG) to be applied to the numerical solution of porous media flows using the Finite Element Method (FEM). The methods described here are general enough to handle with three-dimensional, heterogeneous and anisotropic problems. Examples are shown and results discussed for two-dimensional problems in homogenous and heterogeneous domains with iso- and anisotropic permeability tensors.}\\
\\
\keywords{\textbf{Keywords:} Automated Code Generation, Oil Reservoir, Porous Media Flow, FEM, Multigrid}\\
\end{tabular}

\section{NOMENCLATURE (TO-DO: WHEN PAPER IS DONE, UPDATE THIS LIST)}

@@

\section{INTRODUCTION}
\label{sc:introduction}
Nowadays, the numerical simulation of fluid flow in porous media is fundamental in many engineering fields, such as in the analysis of groundwater flow and contaminant transport \citep{Bear1992}, management of water and heat in polymer membranes for fuel cells \citep{Matamoros2006} or for the simulation of multiphase flows in petroleum reservoirs \citep{Peaceman1977, Aziz1979,Ewing1983,Carvalho2005,Silva2008}.

The scientific modeling of fluid flow in porous media for the petroleum industry has been done since the 1930's, initially using sand-packed models to understand the production of water in oil reservoirs and the reason why the water-oil rate increases with time. In the late 1930's and 1940's some experiments with electrolytic models were used to model the flow in porous media, as it was already recognized the analogy between electric current and the Darcy flow \citep{Peaceman1990}. All these physical analogs models allowed a deeper insight in the phenomena governing this kind of problem, but it was not just until the advent of the electronic computers that engineers were able to simulate real-world problems as, for example, that of oil reservoirs in 2 and 3 dimensions.

Parallel to the development of new hardware, there has been a extensive work on computational methods applied to multiphase flow in porous media. The first method used in large scale for solving partial diffential equations (PDE) numerically and still the standard in the petroleum industry is the finite difference method (FDM) \citep{Peaceman1990}. However, this kind of method presents some disadvantages, as the difficulty to handle complex domains, general boundary conditions or variable material properties \citep{Chen2006}, especially when compared to methods adequate to handle with unstructured meshes that have been applied more recently in this area as the finite volume method (FVM) \citep{Carvalho2005, Cordazzo2006} or the finite element method (FEM) \citep{Chen2006}. For the reasons already cited, the last class of method has been used along this work.

A common aspect among all these computational methods is the considerable amount of time necessary for the development of computer programs that implement formulations for general and/or complex cases. Much of this time is spent coding tasks that are common to almost any numerical simulation software, as for example the matrix assembly, I/O data handling or iterative solutions for linear system of equations. The approach of this work to overcome this difficulty was to use the open source tool \emph{FEniCS/DOLFIN} \citep{Logg2010}, which performs a low-level code (in \emph{C++}),  automatized generation based on some high-level code (in \emph{Python}) input, allowing the developer to focus on developing and testing different mathematical and numerical formulations for the problem of interest using a syntax very similar to the one used for the mathematical description of the problem, instead of spending time writing auxiliary code for ``administrative'' tasks.

Moreover, there is a need to reduce the computing time of the program in order to make the simulations feasible in the time-frame of a project. Normally, the most CPU-time consuming part of a simulation is the solution of the linear equation system resulting from the PDE's discretization with a numerical method \citep{Saad2003}. Meanwhile, several techniques are available to solve iteratively this kind of system, each of them with its own advantages and disadvantages in terms of speed and generality, whereas these mentioned features often lead to distinct directions, that is, a method that is extremely fast for one specific type of problem is not general and vice-versa. In this work, we used one method that is believed to couple almost ideally these two features, namely the Multigrid method \citep{Trottenberg2001, Saad2003, Briggs2000}. Nevertheless, these desired characteristics are not obtained without cost, and in this case the main additional difficulties associated with Multigrid methods are the need of a sequence of progressive coarser meshes whose generation can become an issue if not adequately addressed and the use of operators for data transfer among the successive levels. Following the principle of sparing coding time with general tasks and focusing on the problem of interest, we used in this work the open source library \emph{PyAMG} \citep{Bell2008}, which not only performs in an fully automatic way the matrix-sequence generation given an initial mesh, but also allows the use of many different variants of the Algebraic Multigrid Method (AMG), including the classical Ruge-St\"uben and the Smoothed Aggregation (SA) algorithms \citep{Trottenberg2001}.

\section{MATHEMATICAL FORMULATION} 
In this work, we have adopted the classical mathematical formulation proposed in \cite{Peaceman1977} for the simultaneous flow of two immiscible phases in a saturated porous medium. This approach has been used by many researchers \citep{Ewing1983,Chavent1986,Carvalho2005,Silva2008} and has as one of its main characteristics the manipulation of the mass conservation equation using the Darcy law to form a system of a parabolic-elliptic pressure and a parabolic-hyperbolic saturation equations \citep{Carvalho2005}, as opposed to other formulations where the pressure and saturation fields are solved simultaneously in a system of parabolic PDEs \citep{Aziz1979}. 
%Describe briefly Aziz's formulation in the thesis, for comparison purposes

The segregated formulation of Peaceman allows the use of specialized methods capable of exploring mathematical characteristics for each equation of the resulting system. The coupling of the two fields is achieved through the use of a total velocity term.

In order to use the equations presented in this section, it is necessary to adopt some simplifying assumptions \citep{Carvalho2005,Peaceman1977}:
\begin{itemize}
\item Totally saturated porous medium;
\item Incompressible fluids and rocks;
\item Immiscible fluids;
\item Isothermal flow;
\item Darcy law valid for the fluids velocities.
\end{itemize}

Considering the assumption of totally saturated porous medium and the existence of $n$ phases, the constitutive equation for saturation yields:
\begin{equation} \label{eq:eq_const_sat}
\sum_{\alpha=1}^n S_{\alpha} = 1
\end{equation}

Where $\alpha$ represents each phase (in this work, $\alpha = w$ and $o$ for water and oil, respectively) and $S_{\alpha}$ represents the saturation of the phase $\alpha$.
% Describe concepts as effective porosity and volume, and then give definition of saturation in the thesis.

The generalized Darcy law for each phase velocity reads as:
\begin{equation} \label{eq:eq_darcy_general}
\vec{v}_{\alpha} = - \frac{K_{\alpha}}{\mu_{\alpha}} \nabla\left(p_{\alpha} - \rho_{\alpha}g\nabla z \right)
\end{equation}
with $K_{\alpha}$, $\mu_{\alpha}$, $p_{\alpha}$ and $\rho_{\alpha}$ represent the effective permeability, viscosity, pressure and density of phase $\alpha$, whereas $g$ and $z$ representing the gravity acceleration and the displacement in its direction, respectively. The relation between the effective  and absolute permeability $K$ accounts for the influence of one fluid on the other during the flow and is defined as the relative permeability:
\begin{equation} \label{eq:relat_perm}
k_{r\alpha} = \frac{K_{\alpha}}{K} \leq 1
\end{equation}
%Cite models for relative permeability (Chen,Aziz as general, Brooks-Corey as specific)

The conservation equation for each phase is described as \citep{Peaceman1977}:
\begin{equation} \label{eq:cons_eq}
-\nabla \cdot \left(\rho_{\alpha} \vec{v}_{\alpha}\right) + q_{\alpha} =
\frac{\partial \left(\phi \rho_{\alpha} S_{\alpha}\right)}{\partial t} 
\end{equation}

The terms $q_{\alpha}$, $\phi$ and $t$ represent sources/sinks of phase $\alpha$ (e.g., wells), effective porosity of rock and time, respectively. Further, if we consider two phases, a wetting (water) and a non-wetting (oil), combining the Eqs. (\ref{eq:eq_darcy_general}) and (\ref{eq:cons_eq}), and using the definition given in Eq. (\ref{eq:relat_perm}), we obtain the following system of partial differential equations which solve the two-phase flow problem under the assumptions taken:
\begin{subequations} \label{eq:twophase_system}
  \begin{equation} \label{eq:twophase_systema}
\nabla \cdot \left(\frac{\rho_{w}Kk_{rw}}{\mu_{w}} \nabla\left(p_{w} - \rho_{w}g\nabla z \right)\right) + q_{w} =
\frac{\partial \left(\phi \rho_{w} S_{w}\right)}{\partial t} 
  \end{equation}
  \begin{equation} \label{eq:twophase_systemb}
\nabla \cdot \left(\frac{\rho_{o}Kk_{ro}}{\mu_{o}} \nabla\left(p_{o} - \rho_{o}g\nabla z \right)\right) + q_{o} =
\frac{\partial \left(\phi \rho_{o} S_{o}\right)}{\partial t} 
  \end{equation}
\end{subequations}
%Stress that these equations are more general then the ones necessary for our assumptions and that they will be simplified

In Eqs. (\ref{eq:twophase_systema}) and (\ref{eq:twophase_systemb}), the pressure and saturation fields are coupled in both equations. These resemble the heat conduction equation and are therefore expected to act as essentially parabolic. This assertive is not necessarily true and can be assessed by obtaining a pair of equations that are dependent either on pressure or on saturation. The derivation of these equations is presented in the next subsection.

\subsection{Pressure Equation} \label{sc:press_eq}
The approach used to derive the pressure equation is to eliminate the time derivative of the saturation present in Eq. (\ref{eq:twophase_system}). First, the time derivatives are expanded to obtain:
\begin{subequations} \label{eq:twophase_system2}
  \begin{equation} \label{eq:twophase_system2a}
-\nabla \cdot \left(\rho_{w} \vec{v}_{w}\right) + q_{w} =
\rho_{w}S_{w}\frac{\partial \phi}{\partial t} + \phi S_{w} \frac{d\rho_{w}}{d p_{w}} \frac{\partial p_{w}}{\partial t} + \phi\rho_{w}\frac{\partial S_{w}}{\partial t}
  \end{equation}
  \begin{equation} \label{eq:twophase_system2b}
-\nabla \cdot \left(\rho_{o} \vec{v}_{o}\right) + q_{o} =
\rho_{o}S_{o}\frac{\partial \phi}{\partial t} + \phi S_{o} \frac{d\rho_{o}}{d p_{o}} \frac{\partial p_{o}}{\partial t} + \phi\rho_{o}\frac{\partial S_{o}}{\partial t}
  \end{equation}
\end{subequations}

Then, dividing the first equation by $\rho_{w}$ and the second by $\rho_{o}$, considering the assumptions made previously that fluids and rock are both incompressible and finally adding the resulting equations, we obtain:
% This step mentioned in the previous paragraph can be explained in more 'substeps' in the thesis for clarity
\begin{equation} \label{eq:twophase_oneequation}
-\nabla \cdot \vec{v}_{t} + Q_{t} = \phi \frac{\partial \left(S_{w} + S_{o}\right)}{\partial t}
\end{equation}
where $\vec{v}_{t} = \vec{v}_{w} + \vec{v}_{o}$ is the total velocity of the fluid and $Q_{t} = (q_{w}/\rho_{w}) + (q_{o}/\rho_{o})$ is the total volumetric rate. Moreover, considering Eq. (\ref{eq:eq_const_sat}) and rearranging the terms, we obtain the pressure equation for the two-phase flow in porous media:
\begin{equation} \label{eq:pressure_eq_form1}
\nabla \cdot \vec{v}_{t} = Q_{t}
\end{equation}

In order to present the Eq. (\ref{eq:pressure_eq_form1}) in relation to one single pressure variable, we may define an average pressure by:
\begin{equation} \label{eq:average_press}
p_{avg} = \frac{p_{w} + p_{o}} {2}
\end{equation}

Considering the definition of capillary pressure as $p_{c} = p_{o} - p_{w}$, we may express the individual phase pressures as:
% The capillary pressure must be better explained in the thesis, including some classical models for it
\begin{subequations} \label{eq:phase_pressures_cap}
  \begin{equation} \label{eq:phase_pressures_capa}
p_{w} = p_{avg} - \frac{p_{c}}{2}
  \end{equation}
  \begin{equation} \label{eq:phase_pressures_capb}
p_{o} = p_{avg} + \frac{p_{c}}{2}
  \end{equation}
\end{subequations}

We also define the phase mobilities as the relation between relative permeability and fluid viscosity:
\begin{equation} \label{eq:phase_mob}
\lambda_{\alpha} = \frac{k_{r\alpha}}{\mu_{\alpha}}
\end{equation}

Finally, rewriting Eq. (\ref{eq:pressure_eq_form1}) using average and capillary pressures, we obtain after some rearrangement of the terms:
\begin{equation} \label{eq:pressure_eq_form2}
\nabla \cdot \left(-K \left( \left(\lambda_{w} + \lambda{o} \right)\nabla p_{avg} + \frac{\lambda_{w} - \lambda_{o}}{2}\nabla p_{c} - 
\left(\lambda_{w}\rho_{w} + \lambda_{o}\rho_{o} \right)g\nabla z \right) \right) = Q_{t}
\end{equation}

Thus, it can be clearly observed that Eq. (\ref{eq:pressure_eq_form2}), considering the assumptions made previously, has an elliptic nature. A saturation equation can be deduced using similar algebraic manipulation and will complete the coupled pressure-saturation model for two phase (oil-water) fluid flow in porous media. For the present work we will concentrate on the single-phase flow governed by Eq. (\ref{eq:pressure_eq_form2}) and section \ref{sc:num_form} deals with methods to solve this kind of equation.

\subsection{Saturation Equation}
% Traduzir da tese
\label{sc:eq_sat}
A equação de saturação pode ser deduzida usando uma manipulação algébrica similar à utilizada na seção anterior de modo a completar o modelo de pressão-saturação para um escoamento bifásico água-óleo em meios porosos. Para a dedução desta equação faz-se necessário definir qual das fases (molhante ou não-molhante) servirá de referência. Neste trabalho foi escolhida a fase aquosa seguindo a prática usual da literatura \citep{Peaceman1977, Ewing1983, Carvalho2005, Silva2008}.

Consideraremos a princípio que tenha sido obtida uma solução para um caso mais geral representado pela Eq.~(\ref{eq:pressure_eq_form2}), sendo portanto conhecida $p_{avg}$. Logo, as pressões das fases água $(p_{w})$ e óleo $(p_{o})$ podem ser calculadas a partir da Eq.~(\ref{eq:phase_pressures_cap}) e utilizando a Eq.~(\ref{eq:eq_darcy_general}) podemos calcular as velocidades das duas fases, obtendo:
\begin{subequations} \label{eq:phase_velocities}
  \begin{equation} \label{eq:phase_velocitiesa}
\mathbi{v}_{w} = - \mathbi{K}\lambda_{w} \left(\nabla p_{w} - \rho_{w}g\nabla z \right)
  \end{equation}
  \begin{equation} \label{eq:phase_velocitiesb}
\mathbi{v}_{o} = - \mathbi{K}\lambda_{o} \left(\nabla p_{o} - \rho_{o}g\nabla z \right)
  \end{equation}
\end{subequations}

Multiplicando a Eq.~(\ref{eq:phase_velocitiesa}) por $\lambda_{o}$, a Eq.~(\ref{eq:phase_velocitiesb}) por $\lambda_{w}$, subtraindo uma da outra e utilizando a definição de pressão capilar, pode-se obter:
\begin{equation} \label{eq:phase_velocities2}
- \lambda_{w}\mathbi{v}_{o} + \lambda_{o}\mathbi{v}_{w} = \mathbi{K}\lambda_{o}\lambda_{w}\nabla p_{c} - \mathbi{K}\lambda_{o}\lambda_{w}\left(\rho_{o} - \rho_{w}\right)g\nabla z
\end{equation}

Considerando-se ainda as definições de velocidade e mobilidade total, podemos reescrever a equação anterior como:
\begin{equation} \label{eq:phase_velocities3}
\lambda_{t}\mathbi{v}_{w} = \lambda_{w}\mathbi{v}_{t} + \mathbi{K}\lambda_{o}\lambda_{w}\left(\nabla p_{c} + \left(\rho_{w} - \rho_{o}\right)g\nabla z\right)
\end{equation}

Definindo ainda o fluxo fracional da fase $\alpha$ como:
\begin{equation} \label{eq:fracflux}
f_{\alpha} = \frac{\lambda_{\alpha}}{\lambda_{t}}
\end{equation}%
\nomenclature[aff]{$f$}{Fluxo fracional}%
e utilizando-se a seguinte função da saturação para simplificar a notação:
\begin{equation} \label{eq:derivcapsat}
h_{w} = - \frac{\lambda_{o}\lambda_{w}}{\lambda_{t}} \frac{d p_{c}}{d S_{w}}
\end{equation}
podemos obter uma expressão na qual a velocidade da fase água é expressa em função da velocidade total. Substituindo as Eqs. (\ref{eq:fracflux}) e (\ref{eq:derivcapsat}) na Eq.~(\ref{eq:phase_velocities3}) e inserindo esta na Eq.~(\ref{eq:cons_eq}), podemos escrever a seguinte expressão para a saturação da fase água:
\begin{equation} \label{eq:sateq}
\frac{\partial (\phi\rho_{w}S_{w})}{\partial t} = - \nabla\cdot\left(\rho_{w}\left(f_{w}\mathbi{v}_{t} - \mathbi{K}h_{w}\nabla S_{w} + \mathbi{K}\lambda_{o}f_{w}\left(\rho_{w} - \rho_{o}\right)g\nabla z\right)\right) + q_{w}
\end{equation}

Assumindo as hipóteses já mencionadas de rochas e fluidos incompressíveis podemos representar a expressão anterior por:
\begin{equation} \label{eq:sateq2}
\phi\rho_{w}\frac{\partial S_{w}}{\partial t} = - \rho_{w}\nabla\cdot\left(\left(f_{w}\mathbi{v}_{t} - \mathbi{K}h_{w}\nabla S_{w} + \mathbi{K}\lambda_{o}f_{w}\left(\rho_{w} - \rho_{o}\right)g\nabla z\right)\right) + q_{w}
\end{equation}

Além disso, dividindo a Eq. \ref{eq:sateq2} por $\rho_{w}$ e considerando a definição do termo de taxa volumétrica de injeção/produção da fase água $Q_{w} = q_{w}/\rho_{w}$, obtemos:
\begin{equation} \label{eq:sateq3}
\phi\frac{\partial S_{w}}{\partial t} = - \nabla\cdot\left(\left(f_{w}\mathbi{v}_{t} - \mathbi{K}h_{w}\nabla S_{w} + \mathbi{K}\lambda_{o}f_{w}\left(\rho_{w} - \rho_{o}\right)g\nabla z\right)\right) + Q_{w}
\end{equation}

A Eq.~(\ref{eq:sateq3}) permite obter a saturação da fase água em um meio poroso considerando as hipóteses básicas descritas no início deste capítulo. Esta equação apresenta um comportamento essencialmente parabólico-hiperbólico, a não ser que o termo de capilaridade seja desprezível comparado com os termos advectivos, o que é verdade especialmente para casos como na vizinhança de poços ou em áreas do reservatório onde a velocidade total seja alta \citep{Carvalho2005}. Considerando esta última condição como válida e supondo um escoamento horizontal, obtemos a forma simplificada para a saturação da fase água em um meio poroso:
\begin{equation} \label{eq:sateq4}
\phi\frac{\partial S_{w}}{\partial t} = - \nabla\cdot\left(f_{w}\mathbi{v}_{t}\right) + Q_{w}
\end{equation}

A Eq.~(\ref{eq:sateq4}) apresenta características essencialmente hiperbólicas e não-lineares devido ao termo de fluxo fracional. Um modo de estudar os fenômenos principais associados a esta equação é utilizar modelos que isolem apenas algumas das características de interesse. Sendo assim, serão apresentadas a seguir duas variações das equações apresentadas para a saturação. A primeira delas considera um escoamento unidimensional e sem termo de fonte, onde o principal objetivo é analisar o efeito da não-linearidade decorrente do fluxo fracional. A segunda variação considera uma versão linear da Eq.~(\ref{eq:sateq3}), a qual nada mais é do que a conhecida equação de difusão-convecção e que neste caso governa o escoamento multidimensional de fluidos miscíveis \citep{Peaceman1977, Barbosa2009, Loula1995}.

\subsection{Boundary and Initial Conditions}
% Traduzir da tese
\label{sc:cond_ini_bc}
% Garcia / Chen / Shan / Comentarios Ewing
Antes de especificar a formulação numérica utilizada, algumas definições a respeito das condições iniciais, de contorno e hipóteses simplificadoras têm que ser discutidas para tornar o problema abordado completamente determinado.

Os contornos são descritos como $\Gamma = \partial \Omega$ $= \Gamma_{I} \cup$ $\Gamma_{P} \cup$ $\Gamma_{D} \cup$ $\Gamma_{N}, $%
\nomenclature[e1g]{$\Gamma$}{Contorno do domínio $\Omega$}%
\nomenclature[e1o]{$\Omega$}{Domínio espacial do problema}
onde \citep{Carvalho2005}:
% Talvez mudar a ordem passando a descricao acima para junto da parte de discretizacao,
% comecando portanto apenas com a parte referente a formulacao fraca 'analitica'
\begin{itemize}
\item $\Gamma_{I}$ = Poços injetores;
\item $\Gamma_{P}$ = Poços produtores;
\item $\Gamma_{D}$ = Condição de contorno de Dirichlet (Variável prescrita);
\item $\Gamma_{N}$ = Condição de contorno de Neumman (Fluxo prescrito).
\end{itemize}

Usualmente, os poços são tratados como condições de contorno internas, sendo modelados através de métodos especiais para lidar com a velocidade alta nas adjacências de um poço \citep{Peaceman1977} comparada com o resto do domínio. Apesar disso, foi adotada neste trabalho uma hipótese simplificadora considerando que os poços são termos fonte/sumidouros de produção ou injeção concentrados em um único nó da malha, isto é, uma função delta de Dirac neste ponto.

Neste trabalho, foi feita a consideração usual adotada em problemas de reservatório de petróleo \citep{Peaceman1977, Ewing1983} de condição de fluxo zero em todas as fronteiras exteriores do domínio, isto é:
\begin{equation} \label{eq:noflowbc}
\mathbi{K} \lambda p \cdot \mathbi{n} = 0 \quad \textrm{em $\Gamma_{N}$}
\end{equation}
onde $\lambda$ é a mobibilidade total, ou seja, $\lambda = \lambda_{w} + \lambda_{o}$, $p$ é a pressão média ($p = p_{avg}$) e $\mathbi{n}$%
\nomenclature[bnv]{$\mathbi{n}$}{Vetor normal ao contorno}
 é o vetor unitário normal aos contornos externos.

Logo, assumindo que a transferência de massa entre o reservatório e o meio externo se dará apenas pelos poços injetores e produtores, temos:
\begin{equation} \label{eq:wellbc}
\left. \begin{array}{lll}
\mathbi{v}_{\alpha} \cdot \mathbi{n} = q^{I}_{\alpha} \quad \textrm{ou} \quad p(\mathbi{x},t) = p^{I} \quad \textrm{em $\Gamma_{I}$} \\
\mathbi{v}_{\alpha} \cdot \mathbi{n} = q^{P}_{\alpha} \quad \textrm{ou} \quad p(\mathbi{x},t) = p^{P} \quad \textrm{em $\Gamma_{P}$} \\
\end{array} \right.
\end{equation}
onde $q_{I}_{\alpha}$ e $q_{P}_{\alpha}$ representam as vazões volumétricas da fase $\alpha$ nos poços injetores e produtores, respectivamente.

As condições inicial e de contorno para a equação de saturação são:
\begin{equation} \label{eq:satinibc}
\left. \begin{array}{lll}
S_{w}(\mathbi{x},0) = S^{t=0}_{w} \quad \textrm{em $\Omega$} \\
S_{w}(\mathbi{x},t) = S^{I}_{w} \quad \textrm{em $\Gamma_{I}$} \\
\mathbi{K} \nabla S_{w}(\mathbi{x},t) \cdot \mathbi{n} = 0 \quad \textrm{em $\Gamma_{N}$}\\
\end{array} \right.
\end{equation}

Para a equação da concentração de um soluto em um fluido solvente, ver Eq.~(\ref{eq:misc1}), as condições iniciais e de contorno são definidas de modo similar:
\begin{equation} \label{eq:coninibc}
\left. \begin{array}{lll}
C(\mathbi{x},0) = C^{t=0} \quad \textrm{em $\Omega$} \\
C(\mathbi{x},t) = \hat{C} \quad \textrm{em $\Gamma_{I}$} \\
\mathbi{D} \nabla C(\mathbi{x},t) \cdot \mathbi{n} = 0 \quad \textrm{em $\Gamma_{N}$}\\
\end{array} \right.
\end{equation}

\section{NUMERICAL FORMULATION} \label{sc:num_form}
The basic idea of almost all numerical methods for the solution of PDEs is the discretization and thereby reduction to a finite number of degrees of freedom (DOFs) of a continuum problem, which possesses a infinite number of DOFs. This discrete problem results in an system of equations with finite number of variables, which can normally be solved using an adequate method (see section \ref{sc:multigrid} for details) \citep{Chen2006}.

In the Finite Difference Method (FDM), which is still the standard for numerical reservoir simulations, the derivatives of the original equations are simply replaced by quotients of differences. The values of the variables are defined only in a specified number of points, which can be located either on the corners or in the interior of the cells \citep{Fortuna2000}.

The discretization process within the Finite Element Method (FEM) is different in that it requires a reformulation of the PDEs in an equivalent variational form. The unknown variable, in this case the pressure, is described at each point of the region using an interpolation function. The support points of these functions are the grid nodes, which are connected forming elements.

In this work, we experimented with classical Standard Galerkin Finite Element Method. This methodology is briefly described in the following subsections and references are pointed for further informations.

\subsection{Finite Element Method} \label{sc:fem}
% Garcia / Chen / Shan / Comentarios Ewing
Before specifying the numerical formulation that will be used, some definitions regarding the boundary conditions and simplifying assumptions must be discussed.
In this work, we made the usual consideration adopted in petroleum reservoir problems \citep{Peaceman1977, Ewing1983} of no-flow condition on all the external boundaries of the domain, that is:
\begin{equation} \label{noflowbc}
\lambda K \nabla p \cdot n = 0 \quad \textrm{in $\Gamma_{N}$}
\end{equation}

Where $\lambda$ is the total mobility, that is, $\lambda = \lambda_{w} + \lambda_{o}$, $p$ is the average pressure ($p = p_{avg}$) and $n$ is the unit vector normal to the external boundaries.

The boundaries are described as $\Gamma = \partial \Omega$ $= \Gamma_{I} \cup$ $\Gamma_{P} \cup$ $\Gamma_{D} \cup$ $\Gamma_{N}$, where \citep{Carvalho2005}:
% Talvez mudar a ordem passando a descricao acima para junto da parte de discretizacao,
% comecando portanto apenas com a parte referente a formulacao fraca 'analitica'
\begin{itemize}
\item $\Gamma_{I}$ = Injection Wells;
\item $\Gamma_{P}$ = Production Wells;
\item $\Gamma_{D}$ = Dirichlet Boundary Condition (Prescribed Pressure);
\item $\Gamma_{N}$ = Neumann Boundary Condition (Prescribed Flux).
\end{itemize}

Usually, the wells are treated as internal boundary conditions, being modeled by special methods to deal with the increased velocity in the vicinities of a well \citep{Peaceman1977} compared with the rest of the domain. Despite that, we adopted in this work a simplifying assumption, considering the wells as source/sink terms of injected or produced fluid in a single node of the mesh, that is, a Dirac function in that point. Furthermore, in this work were made the assumptions that the flow is horizontal (no influence of the gravity) and with negligible capillary pressure. This way it was possible to concentrate on the elliptic characteristics of the pressure equation.

In order to solve a problem with the FEM, it is necessary initially to express it in the so-called \emph{weak form}. The first step to obtain this new formulation (also called the variational problem) is to multiply the PDE by a function $w$, called a \emph{test function}, after that the result is integrated over the domain $\Omega$ and finally an integration by parts is performed on terms with second-order derivatives \citep{Hughes2000}. The unknown function (pressure $p$ in this case) is referred to as a \emph{trial function}. Appropriated function spaces must be defined for the test and trial functions.
%Perform the steps above mentioned in a detailed step-by-step presenting the equations
% See Fenics tutorial for reference on that

The variational problem for the pressure equation (Eq. (\ref{eq:pressure_eq_form2})) considering all the mentioned assumptions, has the following form:
\begin{problem} [Continuous]
Find $p$ $\in$ $P$ such that:
\begin{equation} \label{eq:variational}
A(p,w) = f(w) \quad \forall w \in W
\end{equation}
with 
\begin{equation} \label{eq:funcspace1}
P = \{ p \in H^{1}(\Omega), \textrm{$p = p_{D}$ on $\Gamma$} \}  
\end{equation}
\begin{equation} \label{eq:funcspace2}
W = \{ w \in H^{1}(\Omega), \textrm{$w = 0$ on $\Gamma$} \}  
\end{equation}
\begin{equation} \label{eq:bilinearform}
A(p,w) = \int_{\Omega} \lambda K \nabla p \cdot \nabla w d \Omega \quad \forall p,w \in P,W
\end{equation}
\begin{equation} \label{eq:linearform}
f(w) = \int_{\Omega} Q_{t}w d \Omega \quad \forall w \in W
\end{equation}
\end{problem}
where $p_{D}$ is a prescribed value of $p$ on the Dirichlet boundaries and $H^{1}$ is the Sobolev space of functions with square-integrables derivatives of first order. This means that while the PDE solution must lie in a function space with continuous derivatives (\emph{strong form}), the Sobolev space required in the \emph{weak form} allows functions with discontinuous derivatives \citep{Langtangen2010}.
% Explain better about function spaces (Sobolev, L2, etc.). Fenics tutorial good on that
The existence and uniqueness of the solution for this problem are given by the Lax lemma and their full derivation is found in \cite{Garcia1997}.

% Verify the statement about the velocity on well region and the assumptions made for simplification.
For the finite element analysis of this variational problem, we must transform the continuous variational problem defined by Eq. (\ref{eq:variational}) to a discrete one. Consider a domain $\Omega$ which is discretized with $N$ non-overlaping elements $E_{i}$ such that $\Omega_{h} = \bigcup_{i=1}^{N}E_{i}$ and $E_{i} \cap E_{j} = \emptyset, i \not = j$, where the subindex $h$ denotes a mesh size parameter representing the discretized problem.
